---
title: "Assignment7"
author: "Arleen Lindenmeyer"
date: "10/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#install.packages("useful")
library(useful)
library(ggplot2)
```


##Step 1

Run the following code to load some clustering data with 2 features into your session.
```{r}
library(tidyverse)
clus_df <- read_csv("https://raw.githubusercontent.com/vankesteren/dav_practicals/master/12_Unsupervised_learning_Clustering/data/clusterdata.csv")
```

##Step 2

Write a function l2_dist(x, y) that takes in two vectors and outputs the euclidean distance between the two vectors.
```{r}
l2_dist <- function(x, y) sqrt(sum((x - y)^2))
```

Testing the function:
```{r}
x <- c(2, 6, 7, 7, 5, 13, 14, 17, 11, 8)
y <- c(3, 5, 5, 3, 7, 12, 13, 19, 22, 7)
l2_dist(x, y)
```

##Step 3
Program a K-Medioids clustering algorithm called kmedioids. The inputs of this function should be X, a data frame, and K, an integer stating how many clusters you want. The output is at least the cluster assignments (an integer vector). Use your l2_dist function to compute the euclidian distance. Create helpful comments in the code along the way. Apply the kmedioids function on clus_df and visualise your result.

```{r}
#Classifies each point into the cluster whose center it is closest to
cluster_labels <- function (X, centers) {
  df <- data.frame(matrix(ncol = 3, nrow = 0))
  #Calculate distance of points to all center points
  for (c in 1:nrow(X)){
    for (i in 1:nrow(centers)){
        distance = l2_dist(centers[i,-1], X[c,-1])
        df <- rbind(df,(c(centers$X_index[i], X$X_index[c], distance)))
    }
  }

  names <- c("centers_index", "X_index", "distance")
  colnames(df) <- names
  
  #only keep smallest distance of data point to center point > assign data point to center point with smallest distance
  for (m in unique(df$X_index)){
    df_temp <- df[df$X_index == m,]
    df_temp <- df_temp[which.min(df_temp$distance),]
    df <- df[df$X_index != m,]
    df <- rbind(df,df_temp)
  }
  X <- merge(X, df, by.x = "X_index", by.y = "X_index")
  return(X)
}
```

```{r}
#for each point, within one cluster, calculate its distance from all other points within this cluster
calculate_sumdist <- function (X) {
  df2 <- data.frame(matrix(ncol = 3, nrow = 0))
  for (n in 1:nrow(X)){
    sum_dist <- 0
    for (j in 1:nrow(X)){
      distance2 = l2_dist(X[n,2:3], X[j,2:3])
      sum_dist = sum_dist + distance2
    }
    df2 <- rbind(df2, c(X$centers_index[n], X$X_index[n], sum_dist))
  }
  names2 <- c("center_index","X_index", "sum_distance")
  colnames(df2) <- names2
  return(df2)
}
```

```{r}
#big / final function, that uses all other functions above
#takes dataframe X, number of clusters K and maximum number of iterations maxit as input
kmedoids <- function (X, K, maxit) {
  #adding an index column to uniquely identify the data points/ to be used in the functions
  X_with_index <- cbind(X_index = 1:nrow(X), X)
  #Choose k (random) data points as starting points for cluster centers
  centers = X_with_index %>% slice_sample(n = K)
  
  #create before center information to compare current center information to
  before_center_index <- rep(0,K) #for first comparison, to satisfy while    condition
  before_center_coords <- NA
  
  #current center information
  current_center_index <- centers$X_index
  current_center_coords <- X_with_index[X_with_index$X_index %in% current_center_index,]
  
  #keep track of cluster centers
  center_history <- centers$X_index
  
  #counts nr. of iterations, used for maxit
  counter = 0

  #while condition checks if the cluster centers changed (true the vector urrent_center_index is different from before_center_index) and stops the loop after maxit is reached or the cluster centers stayed the same
  while (all(current_center_index != before_center_index) & counter <= maxit) {
    #assignent of cluster labels to the data points
    df_new <- cluster_labels(X_with_index, current_center_coords)
    df_new3  <- data.frame(matrix(ncol = 3, nrow = 0))
  

    for(h in 1:nrow(current_center_coords)){
      center_index = current_center_coords[h,]$X_index
      df_new2 <- df_new %>% filter(centers_index == center_index)
      #calculating sum of distance from one point to all other points            within one cluster
      df_new3 <- rbind(df_new3,calculate_sumdist(df_new2))
      
    }
    
    
    names2 <- c("centers_index", "X_index", "distance")
    colnames(df_new3) <- names2

    #whichever point within one cluster has the smallest sum of distance to     the other points in the cluster is chosen to be the next centroid of       this cluster
    for (m in unique(df_new3$centers_index)){
      df_temp <- df_new3[df_new3$centers_index == m,]
      df_temp <- df_temp[which.min(df_temp$distance),]
      df_new3 <- df_new3[df_new3$centers_index != m,]
      df_new3 <- rbind(df_new3,df_temp)
    }
    
    before_center_index <- current_center_index
    before_center_coords <- current_center_coords

    current_center_coords <- merge(df_new3, X_with_index, by.x = "X_index", by.y = "X_index")
    current_center_coords <- current_center_coords[,c("X_index", "x1", "x2")]
    current_center_index <- current_center_coords$X_index
    
    center_history <- rbind(center_history,current_center_index)

    #counts nr. of iterations, used for maxit
    counter = counter + 1
  }
  
  # create cluster center history
  center_history <- data.frame(center_history)
  center_history$iteration <- 1:nrow(center_history)
  rownames(center_history) <- NULL

  #dataframe with index information, coordinates, label (centers_index) and   if a point is a cluster center
  X_with_index$cluster_center <- "no"
  X_with_index$cluster_center[X_with_index$X_index %in% current_center_coords$X_index] <- "yes"
  X_final <- merge(X_with_index, df_new[,c("X_index","centers_index")], by.x = "X_index", by.y = "X_index")
  X_final$centers_index <- factor(X_final$centers_index)
  
  # create list of outputs (history + data points with assigned clusters/mediods)
  res <- list(clustered_data = X_final, history = center_history)
  
  return(res)
}
```

```{r}
#Apply the kmedioids function on clus_df
res <- kmedoids(clus_df, 3, 10)
res
```


```{r}
# filter out centers for later highlighting in the plot
highlight_centers <- res$clustered_data %>% filter(cluster_center=="yes")

# plot: centers are marked with a cross, the assigment of the points to a cluster is mapped to color
#mapping of the data assigned to clusters
ggplot(res$clustered_data, aes(x=x1, y=x2, color=centers_index)) + 
  geom_point(size=1, alpha = 0.5) +
  #mapping of the centers
  geom_point(data=highlight_centers, 
             aes(x=x1,y=x2), 
             shape='cross',
             size=3) +
  theme_minimal()
```